---
title: "R Notebook"
output: html_notebook
---



```{r, "libraries"}
shhh <- suppressPackageStartupMessages
shhh(library(Matrix))
shhh(library(SingleR))
shhh(library(celldex))
shhh(library(hdf5r))
shhh(library(dplyr))
shhh(library(Seurat))
shhh(library(stringr))
shhh(library(tidyr))
shhh(library(ggplot2))
shhh(library(patchwork))
shhh(library(Seurat))
shhh(library(SingleR))
shhh(library(ggrepel))
shhh(library(qs))
shhh(library(tidytext))
shhh(library(pheatmap))
shhh(library(ComplexHeatmap))
shhh(library(ggpubr))
shhh(library(rstatix))
shhh(library(SingleR))
shhh(library(edgeR))
shhh(library(slingshot))
shhh(library(RColorBrewer))
shhh(library(tradeSeq))
shhh(library(SingleCellExperiment))
```
```{r, "loop functions"}
save.figs <- function(p1,out.dir,base.fn,width,height){
    fn=paste0(base.fn,'.png')
    ggsave(file.path(out.dir,fn),
           plot = p1,
           width = width,
           height = height,
           units = 'in')
    fn=paste0(base.fn,'.pdf')
    ggsave(file.path(out.dir,fn),
           plot = p1,
           width = width,
           height = height,
           units = 'in')
}
volcano.plot <- function(de,p_thresh,log_thresh,title){
    max_abs_x <- max(abs(de$log2FC))
    suppressWarnings({
        p1 <- ggplot(data=de, aes(x=log2FC, y=-log10(p.adj+1e-300), col=diffexpressed, label=delabel)) +
                    geom_point() + 
                    theme_minimal() +
                    geom_text_repel() +
                    scale_color_manual(values=c("blue", "black", "red")) +
                    geom_vline(xintercept=c(-log_thresh, log_thresh), col="red") +
                    geom_hline(yintercept=-log10(p_thresh), col="red") +
                    ggtitle(title) +  xlim(-1.05 * max_abs_x, 1.05 * max_abs_x)
    })

    return(p1)
    }

#HEATMAP.PLOT DOES NOT WORK YET
heatmap.plot <- function(de,sub,title, ds.pt = 100,
                         ndeg_top_bottom=15, 
                          p.thresh=0.05,
                          plot.tag = c('response_status','any_tobacco_smoker')
                          ){

     DefaultAssay(sub) <- "Spatial"
     Idents(object=sub) <- 'cgf.sample.id'
     suppressWarnings({sub <- subset(sub, downsample = ds.pt)})
     DefaultAssay(sub) <- "Spatial"
     exclude.tma <- df[grepl(pattern = 'TMA',x = df$Original.ID),'CGF.Sample.ID']
     annotation_col <- sub@meta.data %>% filter(!(cgf.sample.id %in% exclude.tma)) %>%
       filter(row.names(.) %in% colnames(sub) ) %>%
       select(all_of(plot.tag)) #cgf.sample.id, 
     ndeg = ndeg_top_bottom
     de %>% filter(p_val_adj <  p.thresh) %>% filter(avg_log2FC >0 ) %>%
                 top_n(n = ndeg, wt =  avg_log2FC) -> top
     de %>% filter(p_val_adj <  p.thresh) %>% filter(avg_log2FC <0) %>%
                 top_n(n = ndeg, wt = -avg_log2FC) -> bottom
     temp <- rbind(top,bottom)
     use.genes <- unique(temp$Gene)  
     lognorm.dat <- as.matrix(sub@assaysdata[use.genes,
                                                      row.names(annotation_col),
                                                      drop = FALSE])
     if (length(use.genes) <= 1){
         return(NA)
     }
     ann_colors <- smoker.ann.colors(plot.tag)
     # 5. Plot with pheatmap (row scaling built-in)
     p1 <- pheatmap.wrapper(lognorm.dat, 
                             annotation_col,
                             ann_colors,
                             title)
     return(p1)
}
#part of heatmap.plot, not yet functional
 smoker.ann.colors <- function(plot.tag){
     smoker_colors <- c("yes" = "red", "no" = "gray80")
     response_colors <- c("R" = "green3", "NR" = "black")
     # 4. Define annotation colors
     if (length(plot.tag) == 2){
         ann_colors <- list(
           response_status = response_colors,
           any_tobacco_smoker = smoker_colors
         )
     }else if (plot.tag[1] == 'response_status'){
         ann_colors <- list( response_status = response_colors)
     }else if (plot.tag[1] == 'any_tobacco_smoker'){
         ann_colors <- list(any_tobacco_smoker = smoker_colors)
     }
     return(ann_colors)
 }
 #part of heatmap.plot, not yet functional
 pheatmap.wrapper <- function(lognorm.dat, annotation_col, 
                              ann_colors, title, cluster_rows=TRUE,
                              cluster_cols=TRUE,
                              scale="row"){    
  
    p1<- pheatmap(
       lognorm.dat,
       main = title,
       scale = scale,                 # row-wise z-score normalization
       annotation_col = annotation_col,
       annotation_colors = ann_colors,
       clustering_method = "complete",
       cluster_rows = cluster_rows,
       cluster_cols = cluster_cols,
       color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
       show_colnames = FALSE,
       show_rownames = TRUE
     ) 
     return(p1)
 }
    
```
```{r}
#from brian's deg loop nb
#TIL LOOP SETUP
prj = "HCC-CBS-253-Hillman-LKarapetyan-scRNAseq-revise"
base = file.path('/ix/rbao/Projects', prj)
data = file.path(base,'results','jwork')
scripts= file.path(base,'scripts')
results = file.path(base,'results','jwork')
cell.types <- unique(dat_t@meta.data$general_cell_type)
##TIL LOOP SETUP 2  
use.tags <- c( 'PFS.6month','treatment', 'timepoint') # 'treatment', 'timepoint',
test <- 'MAST'
test.use <- 'MAST'
log.name <- "avg_log2FC"
p.name <- 'p_val_adj'
p = 0.001
log_threshes = c(1, 1.5)
save.filtered <- TRUE
width = 8
height = 6
v=2
suppressWarnings({
    out.gene.lists <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'csv')
    dir.create(out.gene.lists,recursive = TRUE)
    filt.out.gene.lists <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'filtered_csv')
    dir.create(filt.out.gene.lists,recursive = TRUE)
    out.de.volcano <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'volcano')
    dir.create(out.de.volcano,recursive=TRUE)
    out.de.heatmap <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'heatmap')
    dir.create(out.de.heatmap,recursive=TRUE)
})
plot.volc <- TRUE
plot.heat <- FALSE
min.pct <- 0.03
min.cells.feat = 20
tissue.types <- c('TIL')
treatments <- c('rela', 'nivo', 'combo')
timepoints <- c('Baseline','Week4')
pfs = c("Prg6M", "NoPrg6M")
```
```{r}
#TIL LOOP
system.time({
  for (log_thresh in log_threshes){
    for (time in timepoints){
    time.sub <- subset(dat_t, subset = timepoint == time)
      for (tag in use.tags){
        if (tag == 'PFS.6month'){
          idents <- c('NoPrg6M','Prg6M')
        }else if (tag == 'timepoint'){
          idents <- c('Week4','Baseline')
        }else if (tag == 'treatment'){
          idents <- c('rela','nivo','combo')
        }else{
          idents <- c('yes','no')
}
comp.sub <- subset(dat_t, subset = !!sym(tag) %in% idents)
for (tissue in tissue.types){
tissue.sub <- subset(comp.sub, subset = sample_type == tissue)
for (treatment in treatments){
for (cell in cell.types){
if (tag == 'pathological_response_wk4'){
title = sprintf('Respnder vs Non-Responder, %s DEGs',cell)
}else if (tag == 'timepoint'){
title = sprintf('Week4 vs Baseline, %s DEGs',cell)
}else if (tag == 'PFS.6month'){
title = sprintf('NoPrg6m vs Prg6M, %s DEGs',cell)
} else{
title = sprintf('yes vs no, %s DEGs',cell)
}
title = paste(tissue, title)
if (cell != 'all.cells'){
suppressWarnings({sub <- subset(tissue.sub, subset = general_cell_type == cell)})
} else {
sub <- tissue.sub
}
DefaultAssay(sub) <- "RNA"
Idents(sub) <- tag
contrast.str <- paste0(time,'.',treatment,'.',tissue,'.',
tag,'.',idents[1],'.vs.',idents[2])
base.fn <- sprintf('%s.%s.DEG.p%1.3f.log.%1.1f.%s.%d.minpct.v%d',
contrast.str, cell, p,log_thresh, test, min.pct*100,v)
base.fn <- stringr::str_replace_all(base.fn,' ','.')
fn=paste0(base.fn,'.csv')
if (file.exists(file.path(out.gene.lists,fn))){
print(paste('Loading',fn))
de <- read.csv(file.path(out.gene.lists,fn),row.names=1)
}else{
print(paste('Generating',fn))
markers <- FindMarkers(object=sub,
ident.1 = idents[1],
ident.2 = idents[2],
base=2,
min.pct = min.pct,
min.cells.feature = min.cells.feat, #200 for 10% and 10 for 1%
min.cells.group = 50,
test.use = 'MAST')
de <- markers
de$Gene = row.names(markers)
de$log2FC <- de[[log.name]]
max_fc <- 100
de <- de %>%
mutate(
log2FC = case_when(
log2FC == Inf ~ max_fc,
log2FC > max_fc ~ max_fc,
log2FC < -max_fc ~ -max_fc,
log2FC == -Inf ~ -max_fc,
TRUE ~ log2FC
)
) %>%
filter(!is.na(p_val_adj))
de$p.adj <- de[[p.name]]
de$diffexpressed <- "NO"
de$diffexpressed[de$log2FC > log_thresh & de$p.adj < p] <- "UP"
de$diffexpressed[de$log2FC < -log_thresh & de$p.adj < p] <- "DOWN"
de$diffexpressed <- as.character(de$diffexpressed)
#de$Gene = rownames(de)
de$delabel <- NA # Create the column first
de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
#de$delabel<- NA
#de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
de <- de %>% arrange(-log2FC,p_val_adj)
de$contrast <- contrast.str
fn=paste0(base.fn,'.csv')
write.csv(de,file.path(out.gene.lists,fn))
}
if (save.filtered){
filt.de <- de %>% filter(diffexpressed %in% c('UP','DOWN'))
fn = paste0('filt.',base.fn,'.csv')
write.csv(filt.de, file.path(filt.out.gene.lists,fn))
}
if (plot.volc){
p1 <- volcano.plot(de, p, log_thresh, title)
save.figs(p1, out.de.volcano, base.fn, width=width, height=height)
}
if (plot.heat){
if (cell == 'CD8plus.T.cell'){
ds.pt = 10
ndeg = 10
} else {
ds.pt = 200
ndeg = 15
}
plot.tag <- tag
# make_heatmap_bi(de,sub)
#            make_heatmap_bi <- function(de, counts, use_info, output.path, orig.contrast.col=c(), rankby='log2FC',
# p = 0.05,  log_thresh = 1, heatmap_top_plus_minus_n = 25,
# use.p = 'adj.P.Val', require.p.sig.heatmap = TRUE, version = 1,
# heat.colormap = 'warm', save.figs =TRUE, save.name.tag='',cluster_rows = TRUE,
# cluster_cols = TRUE, save_vector_formats = c('.pdf'),
#         added.cols = c(), show_in_line= TRUE){
p2 <- heatmap.plot(de, sub, title, plot.tag = plot.tag,
ndeg_top_bottom = ndeg,  p.thresh=p,
ds.pt =ds.pt)
if (any(!is.na(p2))){
save.figs(p2, out.de.heatmap, base.fn,
width=10, height=5)
}else{
print('Too few genes for heatmap')
}
}
}
}
}
}
}
}
})
```
```{r}
#PBMC LOOP SETUP
dat_p = qread("dat-pbmc-prog-col.qs")
prj = "HCC-CBS-253-Hillman-LKarapetyan-scRNAseq-revise"
base = file.path('/ix/rbao/Projects', prj)
data = file.path(base,'results','jwork')
scripts= file.path(base,'scripts')
results = file.path(base,'results','jwork')
fn <- file.path(results,sprintf('%d_rds_matched_info_pbmc_pfs_03022026_jw.csv',dim(dat_p@meta.data)[1]))
use.tags <- c( 'PFS.6month','timepoint', 'treatment') # 'treatment', 'timepoint',
test <- 'wilcox'
log.name <- "avg_log2FC"
p.name <- 'p_val_adj'
p = 0.001
log_threshes = c(1, 1.5)
save.filtered <- TRUE
width = 8
height = 6
v=1
suppressWarnings({
    out.gene.lists <- file.path(results,sprintf('DEG_contrasts_PBMC_%s_v%d',test,v),'csv')
    dir.create(out.gene.lists,recursive = TRUE)
    filt.out.gene.lists <- file.path(results,sprintf('DEG_contrasts_PBMC_%s_v%d',test,v),'filtered_csv')
    dir.create(filt.out.gene.lists,recursive = TRUE)
    out.de.volcano <- file.path(results,sprintf('DEG_contrasts_PBMC_%s_v%d',test,v),'volcano')
    dir.create(out.de.volcano,recursive=TRUE)
    out.de.heatmap <- file.path(results,sprintf('DEG_contrasts_PBMC_%s_v%d',test,v),'heatmap')
    dir.create(out.de.heatmap,recursive=TRUE)
})
print(fn)
write.csv(dat_p@meta.data, fn)
plot.volc <- TRUE
plot.heat <- FALSE
min.pct <- 0.03
min.cells.feat = 20
tissue.types <- c('PBMC')
treatments <- c('combo', 'nivo','rela')
timepoints <- c('Baseline','Week4') #,'Week16')
```
```{r}
#PBMC loop
system.time({
for (log_thresh in log_threshes){
for (time in timepoints){
time.sub <- subset(dat_p, subset = timepoint == time)
for (tag in use.tags){
if (tag == 'PFS.6month'){
idents <- c('NoPrg6M','Prg6M')
}else if (tag == 'timepoint'){
idents <- c('Week4','Baseline')
}else if (tag == 'treatment'){
idents <- c('combo','rela','nivo')
}else{
idents <- c('yes','no')
}
comp.sub <- subset(dat_p, subset = !!sym(tag) %in% idents)
for (tissue in tissue.types){
tissue.sub <- subset(comp.sub, subset = sample_type == tissue)
for (treatment in treatments){
for (cell in cell.types){
if (tag == 'pathological_response_wk4'){
title = sprintf('Respnder vs Non-Responder, %s DEGs',cell)
}else if (tag == 'timepoint'){
title = sprintf('Week4 vs Baseline, %s DEGs',cell)
}else if (tag == 'PFS.6month'){
title = sprintf('NoPrg6m vs Prg6M, %s DEGs',cell)
} else{
title = sprintf('yes vs no, %s DEGs',cell)
}
title = paste(tissue, title)
if (cell != 'all.cells'){
suppressWarnings({sub <- subset(tissue.sub, subset = general_cell_type == cell)})
} else {
sub <- tissue.sub
}
DefaultAssay(sub) <- "RNA"
Idents(sub) <- tag
contrast.str <- paste0(time,'.',treatment,'.',tissue,'.',
tag,'.',idents[1],'.vs.',idents[2])
base.fn <- sprintf('%s.%s.DEG.p%1.3f.log.%1.1f.%s.%d.minpct.v%d',
contrast.str, cell, p,log_thresh, test, min.pct*100,v)
base.fn <- stringr::str_replace_all(base.fn,' ','.')
fn=paste0(base.fn,'.csv')
if (file.exists(file.path(out.gene.lists,fn))){
print(paste('Loading',fn))
de <- read.csv(file.path(out.gene.lists,fn),row.names=1)
}else{
print(paste('Generating',fn))
markers <- FindMarkers(object=sub,
ident.1 = idents[1],
ident.2 = idents[2],
base=2,
min.pct = min.pct,
min.cells.feature = min.cells.feat, #200 for 10% and 10 for 1%
min.cells.group = 50,
test.use = 'wilcox')# 'negbinom', #'wilcox_limma',#'MAST', #
# latent.vars = c('Patient.Number')) #For models that accept: 'LR', 'negbinom', 'poisson', or 'MAST')
de <- markers
de$Gene = row.names(markers)
de$log2FC <- de[[log.name]]
max_fc <- 100
de <- de %>%
mutate(
log2FC = case_when(
log2FC == Inf ~ max_fc,
log2FC > max_fc ~ max_fc,
log2FC < -max_fc ~ -max_fc,
log2FC == -Inf ~ -max_fc,
TRUE ~ log2FC
)
) %>%
filter(!is.na(p_val_adj))
de$p.adj <- de[[p.name]]
de$diffexpressed <- "NO"
de$diffexpressed[de$log2FC > log_thresh & de$p.adj < p] <- "UP"
de$diffexpressed[de$log2FC < -log_thresh & de$p.adj < p] <- "DOWN"
de$diffexpressed <- as.character(de$diffexpressed)
#de$Gene = rownames(de)
de$delabel <- NA # Create the column first
de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
#de$delabel<- NA
#de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
de <- de %>% arrange(-log2FC,p_val_adj)
de$contrast <- contrast.str
fn=paste0(base.fn,'.csv')
write.csv(de,file.path(out.gene.lists,fn))
}
if (save.filtered){
filt.de <- de %>% filter(diffexpressed %in% c('UP','DOWN'))
fn = paste0('filt.',base.fn,'.csv')
write.csv(filt.de, file.path(filt.out.gene.lists,fn))
}
if (plot.volc){
p1 <- volcano.plot(de, p, log_thresh, title)
save.figs(p1, out.de.volcano, base.fn, width=width, height=height)
}
if (plot.heat){
if (cell == 'CD8plus.T.cell'){
ds.pt = 10
ndeg = 10
} else {
ds.pt = 200
ndeg = 15
}
plot.tag <- tag
# make_heatmap_bi(de,sub)
#            make_heatmap_bi <- function(de, counts, use_info, output.path, orig.contrast.col=c(), rankby='log2FC',
# p = 0.05,  log_thresh = 1, heatmap_top_plus_minus_n = 25,
# use.p = 'adj.P.Val', require.p.sig.heatmap = TRUE, version = 1,
# heat.colormap = 'warm', save.figs =TRUE, save.name.tag='',cluster_rows = TRUE,
# cluster_cols = TRUE, save_vector_formats = c('.pdf'),
#         added.cols = c(), show_in_line= TRUE){
p2 <- heatmap.plot(de, sub, title, plot.tag = plot.tag,
ndeg_top_bottom = ndeg,  p.thresh=p,
ds.pt =ds.pt)
if (any(!is.na(p2))){
save.figs(p2, out.de.heatmap, base.fn,
width=10, height=5)
}else{
print('Too few genes for heatmap')
}
}
}
}
}
}
}
}
})
```

```{r}
#FIXED BY CLAUDE
#TIL LOOP SETUP
prj = "HCC-CBS-253-Hillman-LKarapetyan-scRNAseq-revise"
base = file.path('/ix/rbao/Projects', prj)
data = file.path(base,'results','jwork')
scripts= file.path(base,'scripts')
results = file.path(base,'results','jwork')
cell.types <- unique(dat_t@meta.data$general_cell_type)
##TIL LOOP SETUP 2  
use.tags <- c( 'PFS.6month','treatment', 'timepoint') # 'treatment', 'timepoint',
test <- 'MAST'
log.name <- "avg_log2FC"
p.name <- 'p_val_adj'
p = 0.001
log_threshes = c(1, 1.5)
save.filtered <- TRUE
width = 8
height = 6
v=4
suppressWarnings({
    out.gene.lists <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'csv')
    dir.create(out.gene.lists,recursive = TRUE)
    filt.out.gene.lists <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'filtered_csv')
    dir.create(filt.out.gene.lists,recursive = TRUE)
    out.de.volcano <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'volcano')
    dir.create(out.de.volcano,recursive=TRUE)
    out.de.heatmap <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'heatmap')
    dir.create(out.de.heatmap,recursive=TRUE)
})
plot.volc <- TRUE
plot.heat <- TRUE
min.pct <- 0.03
min.cells.feat = 20
tissue.types <- c('TIL')
treatments <- c('rela', 'nivo', 'combo')
timepoints <- c('Baseline','Week4')
PFS.6month <- c('Prg6M','NoPrg6M')
```
```{r}
#TIL LOOP
system.time({
for (log_thresh in log_threshes){
  for (time in timepoints){
    time.sub <- subset(dat_t, subset = timepoint == time)      # ✓ fixed
    for (tag in use.tags){
      if (tag == 'PFS.6month'){
        idents <- c('NoPrg6M','Prg6M')
      } else if (tag == 'timepoint'){
        idents <- c('Week4','Baseline')
      } else if (tag == 'treatment'){
        contrast.pairs = list(
          c('rela', 'nivo'),
          c('rela', 'combo'),
          c('nivo', 'combo'))
      } else {
        contrast.pairs = list(idents[1:2]) }
      } else {
        idents <- c('yes','no')
      }
      comp.sub <- subset(time.sub, subset = !!sym(tag) %in% idents)  # ✓ fixed
      for (tissue in tissue.types){
        tissue.sub <- subset(comp.sub, subset = sample_type == tissue)
        for (trt in treatments){                                       # ✓ renamed
          treat.sub <- subset(tissue.sub, subset = treatment == trt)  # ✓ added
          for (cell in cell.types){
            if (tag == 'PFS.6month'){
              title = sprintf('NoPrg6m vs Prg6M, %s DEGs', cell)
            } else if (tag == 'timepoint'){
              title = sprintf('Week4 vs Baseline, %s DEGs', cell)
            } else if (tag == 'treatment'){
              title = sprintf('rela vs nivo vs combo, %s DEGs', cell) # ✓ fixed
            } else {
              title = sprintf('yes vs no, %s DEGs', cell)             # ✓ fixed brace
            }
            title = paste(tissue, title)
            if (cell != 'all.cells'){
              sub <- subset(treat.sub, subset = general_cell_type == cell) # ✓ uses treat.sub
            } else {
              sub <- treat.sub                                              # ✓ uses treat.sub
            }
            DefaultAssay(sub) <- "RNA"
            Idents(sub) <- tag
            contrast.str <- paste0(time,'.',trt,'.',tissue,'.',            # ✓ uses trt
                                   tag,'.',idents[1],'.vs.',idents[2])
base.fn <- sprintf('%s.%s.DEG.p%1.3f.log.%1.1f.%s.%d.minpct.v%d',
contrast.str, cell, p,log_thresh, test, min.pct*100,v)
base.fn <- stringr::str_replace_all(base.fn,' ','.')
fn=paste0(base.fn,'.csv')
if (file.exists(file.path(out.gene.lists,fn))){
print(paste('Loading',fn))
de <- read.csv(file.path(out.gene.lists,fn),row.names=1)
}else{
print(paste('Generating',fn))
markers <- FindMarkers(object=sub,
ident.1 = idents[1],
ident.2 = idents[2],
base=2,
min.pct = min.pct,
min.cells.feature = min.cells.feat, #200 for 10% and 10 for 1%
min.cells.group = 50,
test.use = 'MAST')# 'negbinom', #'wilcox_limma',#'MAST', #
# latent.vars = c('Patient.Number')) #For models that accept: 'LR', 'negbinom', 'poisson', or 'MAST')
de <- markers
de$Gene = row.names(markers)
de$log2FC <- de[[log.name]]
max_fc <- 100
de <- de %>%
mutate(
log2FC = case_when(
log2FC == Inf ~ max_fc,
log2FC > max_fc ~ max_fc,
log2FC < -max_fc ~ -max_fc,
log2FC == -Inf ~ -max_fc,
TRUE ~ log2FC
)
) %>%
filter(!is.na(p_val_adj))
de$p.adj <- de[[p.name]]
de$diffexpressed <- "NO"
de$diffexpressed[de$log2FC > log_thresh & de$p.adj < p] <- "UP"
de$diffexpressed[de$log2FC < -log_thresh & de$p.adj < p] <- "DOWN"
de$diffexpressed <- as.character(de$diffexpressed)
#de$Gene = rownames(de)
de$delabel <- NA # Create the column first
de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
#de$delabel<- NA
#de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
de <- de %>% arrange(-log2FC,p_val_adj)
de$contrast <- contrast.str
fn=paste0(base.fn,'.csv')
write.csv(de,file.path(out.gene.lists,fn))
}
if (save.filtered){
filt.de <- de %>% filter(diffexpressed %in% c('UP','DOWN'))
fn = paste0('filt.',base.fn,'.csv')
write.csv(filt.de, file.path(filt.out.gene.lists,fn))
}
if (plot.volc){
p1 <- volcano.plot(de, p, log_thresh, title)
save.figs(p1, out.de.volcano, base.fn, width=width, height=height)
}
if (plot.heat){
if (cell == cell){
ds.pt = 10
ndeg = 10
} else {
ds.pt = 200
ndeg = 15
}
plot.tag <- tag
# make_heatmap_bi(de,sub)
#            make_heatmap_bi <- function(de, counts, use_info, output.path, orig.contrast.col=c(), rankby='log2FC',
# p = 0.05,  log_thresh = 1, heatmap_top_plus_minus_n = 25,
# use.p = 'adj.P.Val', require.p.sig.heatmap = TRUE, version = 1,
# heat.colormap = 'warm', save.figs =TRUE, save.name.tag='',cluster_rows = TRUE,
# cluster_cols = TRUE, save_vector_formats = c('.pdf'),
#         added.cols = c(), show_in_line= TRUE){
p2 <- heatmap.plot(de, sub, title, plot.tag = plot.tag,
ndeg_top_bottom = ndeg,  p.thresh=p,
ds.pt =ds.pt)
if (any(!is.na(p2))){
save.figs(p2, out.de.heatmap, base.fn,
width=10, height=5)
}else{
print('Too few genes for heatmap')
}
}
}
}
}
}
}
}
})

```
```{r}
#claude fix 2
#TIL LOOP
system.time({
for (log_thresh in log_threshes){
for (time in timepoints){
  time.sub <- subset(dat_t, subset = timepoint == time)
  for (tag in use.tags){
    if (tag == 'PFS.6month'){
      idents <- c('NoPrg6M','Prg6M')
    } else if (tag == 'timepoint'){
      idents <- c('Week4','Baseline')
    } else if (tag == 'treatment'){
      idents <- c('combo','rela','nivo')
    } else {
      idents <- c('yes','no')
    }
    
    # Define pairwise contrasts
    if (tag == 'treatment'){
      contrast.pairs <- list(c('combo','rela'), c('combo','nivo'), c('rela','nivo'))
    } else {
      contrast.pairs <- list(idents[1:2])
    }
    
    comp.sub <- subset(time.sub, subset = !!sym(tag) %in% idents)  # fixed: was dat_t
    
    for (tissue in tissue.types){
      tissue.sub <- subset(comp.sub, subset = sample_type == tissue)
      
      for (trt in treatments){
        # Don't subset by treatment when treatment IS the comparison tag
        if (tag == 'treatment'){
          if (trt != treatments[1]) next  # run only once for treatment comparisons
          treat.sub <- tissue.sub         # keep all treatments present
        } else {
          treat.sub <- subset(tissue.sub, subset = treatment == trt)
        }
        
        for (cell in cell.types){
          # Guard: skip if no cells found
          if (ncol(treat.sub) == 0){
            message('Skipping: no cells for trt=', trt, ' tissue=', tissue, ' tag=', tag)
            next
          }
          
          if (cell != 'all.cells'){
            sub <- suppressWarnings(subset(treat.sub, subset = general_cell_type == cell))
          } else {
            sub <- treat.sub
          }
          
          # Guard: skip if subset is empty
          if (ncol(sub) == 0){
            message('Skipping: no cells for cell=', cell)
            next
          }
          
          DefaultAssay(sub) <- "RNA"
          Idents(sub) <- tag
          
          # Title
          if (tag == 'PFS.6month'){
            title = sprintf('NoPrg6m vs Prg6M, %s DEGs', cell)
          } else if (tag == 'timepoint'){
            title = sprintf('Week4 vs Baseline, %s DEGs', cell)
          } else if (tag == 'treatment'){
            title = sprintf('Treatment comparisons, %s DEGs', cell)
          } else {
            title = sprintf('yes vs no, %s DEGs', cell)
          }
          title = paste(tissue, title)
          
          # Loop over pairwise contrasts
          for (pair in contrast.pairs){
            ident.1 <- pair[1]
            ident.2 <- pair[2]
            
            # Guard: check both idents exist in this subset
            available.idents <- unique(as.character(sub[[tag]][,1]))
            if (!ident.1 %in% available.idents | !ident.2 %in% available.idents){
              message('Skipping: missing ident in subset. Need: ', ident.1, ' and ', 
                      ident.2, '. Found: ', paste(available.idents, collapse=', '))
              next
            }
            
            # Update title for pairwise treatment contrasts
            if (tag == 'treatment'){
              title = sprintf('%s vs %s, %s DEGs', ident.1, ident.2, cell)
              title = paste(tissue, title)
            }
            
            contrast.str <- paste0(time,'.',trt,'.',tissue,'.',
                                   tag,'.',ident.1,'.vs.',ident.2)  # fixed: uses trt
            base.fn <- sprintf('%s.%s.DEG.p%1.3f.log.%1.1f.%s.%d.minpct.v%d',
                               contrast.str, cell, p, log_thresh, test, min.pct*100, v)
            base.fn <- stringr::str_replace_all(base.fn, ' ', '.')
            fn <- paste0(base.fn, '.csv')
            
            if (file.exists(file.path(out.gene.lists, fn))){
              print(paste('Loading', fn))
              de <- read.csv(file.path(out.gene.lists, fn), row.names=1)
            } else {
              print(paste('Generating', fn))
              markers <- tryCatch({
                FindMarkers(object = sub,
                            ident.1 = ident.1,
                            ident.2 = ident.2,
                            base = 2,
                            min.pct = min.pct,
                            min.cells.feature = min.cells.feat,
                            min.cells.group = 50,
                            test.use = 'MAST')
              }, error = function(e){
                message('FindMarkers failed for ', contrast.str, ': ', e$message)
                return(NULL)
              })
              
              if (is.null(markers) || nrow(markers) == 0){
                message('No markers found for ', contrast.str, ', skipping')
                next
              }
              
              de <- markers
              de$Gene <- row.names(markers)
              de$log2FC <- de[[log.name]]
              max_fc <- 100
              de <- de %>%
                mutate(
                  log2FC = case_when(
                    log2FC == Inf   ~ max_fc,
                    log2FC > max_fc ~ max_fc,
                    log2FC < -max_fc ~ -max_fc,
                    log2FC == -Inf  ~ -max_fc,
                    TRUE            ~ log2FC
                  )
                ) %>%
                filter(!is.na(p_val_adj))
              
              de$p.adj <- de[[p.name]]
              de$diffexpressed <- "NO"
              de$diffexpressed[de$log2FC > log_thresh  & de$p.adj < p] <- "UP"
              de$diffexpressed[de$log2FC < -log_thresh & de$p.adj < p] <- "DOWN"
              de$diffexpressed <- as.character(de$diffexpressed)
              de$delabel <- NA
              de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
              de <- de %>% arrange(-log2FC, p_val_adj)
              de$contrast <- contrast.str
              
              write.csv(de, file.path(out.gene.lists, fn))
            }
            
            if (save.filtered){
              filt.de <- de %>% filter(diffexpressed %in% c('UP','DOWN'))
              fn.filt <- paste0('filt.', base.fn, '.csv')
              write.csv(filt.de, file.path(filt.out.gene.lists, fn.filt))
            }
            
            if (plot.volc){
              p1 <- volcano.plot(de, p, log_thresh, title)
              save.figs(p1, out.de.volcano, base.fn, width=width, height=height)
            }
            
            if (plot.heat){
              if (cell == 'all.cells'){   # fixed: was cell = cell
                ds.pt <- 10
                ndeg  <- 10
              } else {
                ds.pt <- 200
                ndeg  <- 15
              }
              plot.tag <- tag
              p2 <- heatmap.plot(de, sub, title, plot.tag = plot.tag,
                                 ndeg_top_bottom = ndeg, p.thresh = p,
                                 ds.pt = ds.pt)
              if (any(!is.na(p2))){
                save.figs(p2, out.de.heatmap, base.fn, width=10, height=5)
              } else {
                print('Too few genes for heatmap')
              }
            }
            
          } # end pair loop
        } # end cell loop
      } # end trt loop
    } # end tissue loop
  } # end tag loop
} # end time loop
} # end log_thresh loop
}) # end system.time

```
```{r}
#claude til + new heatmap plot func
#TIL claude fix 4
#TIL LOOP SETUP
prj = "HCC-CBS-253-Hillman-LKarapetyan-scRNAseq-revise"
base = file.path('/ix/rbao/Projects', prj)
data = file.path(base,'results','jwork')
scripts= file.path(base,'scripts')
results = file.path(base,'results','jwork')
cell.types <- unique(dat_t@meta.data$general_cell_type)
##TIL LOOP SETUP 2  
use.tags <- c( 'PFS.6month','treatment', 'timepoint') # 'treatment', 'timepoint',
test <- 'MAST'
log.name <- "avg_log2FC"
p.name <- 'p_val_adj'
p = 0.001
log_threshes = c(1, 1.5)
save.filtered <- TRUE
width = 8
height = 6
v=3
suppressWarnings({
    out.gene.lists <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'csv')
    dir.create(out.gene.lists,recursive = TRUE)
    filt.out.gene.lists <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'filtered_csv')
    dir.create(filt.out.gene.lists,recursive = TRUE)
    out.de.volcano <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'volcano')
    dir.create(out.de.volcano,recursive=TRUE)
    out.de.heatmap <- file.path(results,sprintf('DEG_contrasts_TIL_%s_v%d',test,v),'heatmap')
    dir.create(out.de.heatmap,recursive=TRUE)
})
plot.volc <- TRUE
plot.heat <- TRUE
min.pct <- 0.03
min.cells.feat = 20
tissue.types <- c('TIL')
treatments <- c('rela', 'nivo', 'combo')
timepoints <- c('Baseline','Week4')
PFS.6month <- c('Prg6M','NoPrg6M')

#TIL new loop with heatmap
system.time({
for (log_thresh in log_threshes){
for (time in timepoints){
for (tag in use.tags){

    # --- Define idents ---
    if (tag == 'PFS.6month'){
      idents <- c('NoPrg6M','Prg6M')
    } else if (tag == 'timepoint'){
      idents <- c('Week4','Baseline')
    } else if (tag == 'treatment'){
      idents <- c('combo','rela','nivo')
    } else {
      idents <- c('yes','no')
    }

    # --- Define pairwise contrasts ---
    if (tag == 'treatment'){
      contrast.pairs <- list(c('combo','rela'), c('combo','nivo'), c('rela','nivo'))
    } else {
      contrast.pairs <- list(idents[1:2])
    }

    # --- Timepoint subsetting: skip redundant iterations when tag == 'timepoint' ---
    if (tag == 'timepoint'){
      if (time != timepoints[1]) next  # run only once across time loop
      time.sub <- dat_t                # keep all timepoints for comparison
    } else {
      time.sub <- subset(dat_t, subset = timepoint == time)
    }

    comp.sub <- subset(time.sub, subset = !!sym(tag) %in% idents)

    for (tissue in tissue.types){
      tissue.sub <- subset(comp.sub, subset = sample_type == tissue)

      for (trt in treatments){

        # --- Treatment subsetting: skip redundant iterations when tag == 'treatment' ---
        if (tag == 'treatment'){
          if (trt != treatments[1]) next
          treat.sub <- tissue.sub
        } else {
          treat.sub <- subset(tissue.sub, subset = treatment == trt)
        }

        for (cell in cell.types){

          if (ncol(treat.sub) == 0){
            message('Skipping: no cells for trt=', trt, ' tissue=', tissue, ' tag=', tag)
            next
          }

          if (cell != 'all.cells'){
            sub <- suppressWarnings(subset(treat.sub, subset = general_cell_type == cell))
          } else {
            sub <- treat.sub
          }

          if (ncol(sub) == 0){
            message('Skipping: no cells for cell=', cell)
            next
          }

          DefaultAssay(sub) <- "RNA"
          Idents(sub) <- tag

          # --- Title ---
          if (tag == 'PFS.6month'){
            title = sprintf('NoPrg6m vs Prg6M, %s DEGs', cell)
          } else if (tag == 'timepoint'){
            title = sprintf('Week4 vs Baseline, %s DEGs', cell)
          } else if (tag == 'treatment'){
            title = sprintf('Treatment comparisons, %s DEGs', cell)
          } else {
            title = sprintf('yes vs no, %s DEGs', cell)
          }
          title = paste(tissue, title)

          for (pair in contrast.pairs){
            ident.1 <- pair[1]
            ident.2 <- pair[2]

            # --- Guard: both idents must exist in subset ---
            available.idents <- unique(as.character(sub[[tag]][,1]))
            if (!ident.1 %in% available.idents || !ident.2 %in% available.idents){
              message('Skipping: missing ident. Need: ', ident.1, ' and ', ident.2,
                      '. Found: ', paste(available.idents, collapse=', '))
              next
            }

            if (tag == 'treatment'){
              title = paste(tissue, sprintf('%s vs %s, %s DEGs', ident.1, ident.2, cell))
            }

            # --- For timepoint tag, use 'all' as time label in filename ---
            time.label <- if (tag == 'timepoint') 'alltime' else time
            trt.label  <- if (tag == 'treatment') 'alltrt'  else trt

            contrast.str <- paste0(time.label,'.',trt.label,'.',tissue,'.',
                                   tag,'.',ident.1,'.vs.',ident.2)
            base.fn <- sprintf('%s.%s.DEG.p%1.3f.log.%1.1f.%s.%d.minpct.v%d',
                               contrast.str, cell, p, log_thresh, test, min.pct*100, v)
            base.fn <- stringr::str_replace_all(base.fn, ' ', '.')
            fn <- paste0(base.fn, '.csv')

            if (file.exists(file.path(out.gene.lists, fn))){
              print(paste('Loading', fn))
              de <- read.csv(file.path(out.gene.lists, fn), row.names=1)
            } else {
              print(paste('Generating', fn))
              markers <- tryCatch({
                FindMarkers(object = sub,
                            ident.1 = ident.1,
                            ident.2 = ident.2,
                            base = 2,
                            min.pct = min.pct,
                            min.cells.feature = min.cells.feat,
                            min.cells.group = 50,
                            test.use = 'MAST')
              }, error = function(e){
                message('FindMarkers failed for ', contrast.str, ': ', e$message)
                return(NULL)
              })

              if (is.null(markers) || nrow(markers) == 0){
                message('No markers found for ', contrast.str, ', skipping')
                next
              }

              de <- markers
              de$Gene <- row.names(markers)
              de$log2FC <- de[[log.name]]
              max_fc <- 100
              de <- de %>%
                mutate(
                  log2FC = case_when(
                    log2FC == Inf    ~ max_fc,
                    log2FC > max_fc  ~ max_fc,
                    log2FC < -max_fc ~ -max_fc,
                    log2FC == -Inf   ~ -max_fc,
                    TRUE             ~ log2FC
                  )
                ) %>%
                filter(!is.na(p_val_adj))

              de$p.adj <- de[[p.name]]
              de$diffexpressed <- "NO"
              de$diffexpressed[de$log2FC > log_thresh  & de$p.adj < p] <- "UP"
              de$diffexpressed[de$log2FC < -log_thresh & de$p.adj < p] <- "DOWN"
              de$diffexpressed <- as.character(de$diffexpressed)
              de$delabel <- NA
              de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
              de <- de %>% arrange(-log2FC, p_val_adj)
              de$contrast <- contrast.str

              write.csv(de, file.path(out.gene.lists, fn))
            }

            if (save.filtered){
              filt.de <- de %>% filter(diffexpressed %in% c('UP','DOWN'))
              fn.filt <- paste0('filt.', base.fn, '.csv')
              write.csv(filt.de, file.path(filt.out.gene.lists, fn.filt))
            }

            if (plot.volc){
              p1 <- volcano.plot(de, p, log_thresh, title)
              save.figs(p1, out.de.volcano, base.fn, width=width, height=height)
            }

            if (plot.heat){
              if (cell == 'all.cells'){
                ds.pt <- 10
                ndeg  <- 10
              } else {
                ds.pt <- 200
                ndeg  <- 15
              }
              plot.tag <- tag
              p2 <- heatmap.plot(de, sub, title, plot.tag = plot.tag,
                                 ndeg_top_bottom = ndeg, p.thresh = p,
                                 ds.pt = ds.pt)
              if (any(!is.na(p2))){
                save.figs(p2, out.de.heatmap, base.fn, width=10, height=5)
              } else {
                print('Too few genes for heatmap')
              }
            }

          } # end pair loop
        } # end cell loop
      } # end trt loop
    } # end tissue loop
} # end tag loop  ← moved inside time loop so tag is defined before time.sub logic
} # end time loop
} # end log_thresh loop
}) # end system.time
```
```{r}
#PBMC claude fix 4
#PBMC LOOP SETUP
prj = "HCC-CBS-253-Hillman-LKarapetyan-scRNAseq-revise"
base = file.path('/ix/rbao/Projects', prj)
data = file.path(base,'results','jwork')
scripts= file.path(base,'scripts')
results = file.path(base,'results','jwork')
cell.types <- unique(dat_p@meta.data$general_cell_type)
##PBMC LOOP SETUP 2  
use.tags <- c( 'PFS.6month','treatment', 'timepoint') # 'treatment', 'timepoint',
test <- 'MAST'
log.name <- "avg_log2FC"
p.name <- 'p_val_adj'
p = 0.001
log_threshes = c(1, 1.5)
save.filtered <- TRUE
width = 8
height = 6
v=4
suppressWarnings({
    out.gene.lists <- file.path(results,sprintf('DEG_contrasts_PBMC_%s_v%d',test,v),'csv')
    dir.create(out.gene.lists,recursive = TRUE)
    filt.out.gene.lists <- file.path(results,sprintf('DEG_contrasts_PBMC_%s_v%d',test,v),'filtered_csv')
    dir.create(filt.out.gene.lists,recursive = TRUE)
    out.de.volcano <- file.path(results,sprintf('DEG_contrasts_PBMC_%s_v%d',test,v),'volcano')
    dir.create(out.de.volcano,recursive=TRUE)
    out.de.heatmap <- file.path(results,sprintf('DEG_contrasts_PBMC_%s_v%d',test,v),'heatmap')
    dir.create(out.de.heatmap,recursive=TRUE)
})
plot.volc <- TRUE
plot.heat <- TRUE
min.pct <- 0.03
min.cells.feat = 20
tissue.types <- c('PBMC')
treatments <- c('rela', 'nivo', 'combo')
timepoints <- c('Baseline','Week4')
PFS.6month <- c('Prg6M','NoPrg6M')
```
```{r}
# Run outside heatmap.plot to identify the correct assay
cat("Assay names:", paste(Assays(sub), collapse=", "), "\n")
cat("RNA nrow:", nrow(sub[["RNA"]]), "\n")
cat("First 5 RNA rownames:", paste(head(rownames(sub[["RNA"]]), 5), collapse=", "), "\n")

# Check if genes are in a different assay
for (assay_name in Assays(sub)) {
  n_match <- sum(use.genes %in% rownames(sub[[assay_name]]))
  cat(assay_name, "matches:", n_match, "/", length(use.genes), "\n")
}
```
```{r}
#PBMC LOOP
system.time({
for (log_thresh in log_threshes){
for (time in timepoints){
for (tag in use.tags){

    # --- Define idents ---
    if (tag == 'PFS.6month'){
      idents <- c('NoPrg6M','Prg6M')
    } else if (tag == 'timepoint'){
      idents <- c('Week4','Baseline')
    } else if (tag == 'treatment'){
      idents <- c('combo','rela','nivo')
    } else {
      idents <- c('yes','no')
    }

    # --- Define pairwise contrasts ---
    if (tag == 'treatment'){
      contrast.pairs <- list(c('combo','rela'), c('combo','nivo'), c('rela','nivo'))
    } else {
      contrast.pairs <- list(idents[1:2])
    }

    # --- Timepoint subsetting: skip redundant iterations when tag == 'timepoint' ---
    if (tag == 'timepoint'){
      if (time != timepoints[1]) next  # run only once across time loop
      time.sub <- dat_p                # keep all timepoints for comparison
    } else {
      time.sub <- subset(dat_p, subset = timepoint == time)
    }

    comp.sub <- subset(time.sub, subset = !!sym(tag) %in% idents)

    for (tissue in tissue.types){
      tissue.sub <- subset(comp.sub, subset = sample_type == tissue)

      for (trt in treatments){

        # --- Treatment subsetting: skip redundant iterations when tag == 'treatment' ---
        if (tag == 'treatment'){
          if (trt != treatments[1]) next
          treat.sub <- tissue.sub
        } else {
          treat.sub <- subset(tissue.sub, subset = treatment == trt)
        }

        for (cell in cell.types){

          if (ncol(treat.sub) == 0){
            message('Skipping: no cells for trt=', trt, ' tissue=', tissue, ' tag=', tag)
            next
          }

          if (cell != 'all.cells'){
            sub <- suppressWarnings(subset(treat.sub, subset = general_cell_type == cell))
          } else {
            sub <- treat.sub
          }

          if (ncol(sub) == 0){
            message('Skipping: no cells for cell=', cell)
            next
          }

          DefaultAssay(sub) <- "RNA"
          Idents(sub) <- tag

          # --- Title ---
          if (tag == 'PFS.6month'){
            title = sprintf('NoPrg6m vs Prg6M, %s DEGs', cell)
          } else if (tag == 'timepoint'){
            title = sprintf('Week4 vs Baseline, %s DEGs', cell)
          } else if (tag == 'treatment'){
            title = sprintf('Treatment comparisons, %s DEGs', cell)
          } else {
            title = sprintf('yes vs no, %s DEGs', cell)
          }
          title = paste(tissue, title)

          for (pair in contrast.pairs){
            ident.1 <- pair[1]
            ident.2 <- pair[2]

            # --- Guard: both idents must exist in subset ---
            available.idents <- unique(as.character(sub[[tag]][,1]))
            if (!ident.1 %in% available.idents || !ident.2 %in% available.idents){
              message('Skipping: missing ident. Need: ', ident.1, ' and ', ident.2,
                      '. Found: ', paste(available.idents, collapse=', '))
              next
            }

            if (tag == 'treatment'){
              title = paste(tissue, sprintf('%s vs %s, %s DEGs', ident.1, ident.2, cell))
            }

            # --- For timepoint tag, use 'all' as time label in filename ---
            time.label <- if (tag == 'timepoint') 'alltime' else time
            trt.label  <- if (tag == 'treatment') 'alltrt'  else trt

            contrast.str <- paste0(time.label,'.',trt.label,'.',tissue,'.',
                                   tag,'.',ident.1,'.vs.',ident.2)
            base.fn <- sprintf('%s.%s.DEG.p%1.3f.log.%1.1f.%s.%d.minpct.v%d',
                               contrast.str, cell, p, log_thresh, test, min.pct*100, v)
            base.fn <- stringr::str_replace_all(base.fn, ' ', '.')
            fn <- paste0(base.fn, '.csv')

            if (file.exists(file.path(out.gene.lists, fn))){
              print(paste('Loading', fn))
              de <- read.csv(file.path(out.gene.lists, fn), row.names=1)
            } else {
              print(paste('Generating', fn))
              markers <- tryCatch({
                FindMarkers(object = sub,
                            ident.1 = ident.1,
                            ident.2 = ident.2,
                            base = 2,
                            min.pct = min.pct,
                            min.cells.feature = min.cells.feat,
                            min.cells.group = 50,
                            test.use = 'MAST')
              }, error = function(e){
                message('FindMarkers failed for ', contrast.str, ': ', e$message)
                return(NULL)
              })

              if (is.null(markers) || nrow(markers) == 0){
                message('No markers found for ', contrast.str, ', skipping')
                next
              }

              de <- markers
              de$Gene <- row.names(markers)
              de$log2FC <- de[[log.name]]
              max_fc <- 100
              de <- de %>%
                mutate(
                  log2FC = case_when(
                    log2FC == Inf    ~ max_fc,
                    log2FC > max_fc  ~ max_fc,
                    log2FC < -max_fc ~ -max_fc,
                    log2FC == -Inf   ~ -max_fc,
                    TRUE             ~ log2FC
                  )
                ) %>%
                filter(!is.na(p_val_adj))

              de$p.adj <- de[[p.name]]
              de$diffexpressed <- "NO"
              de$diffexpressed[de$log2FC > log_thresh  & de$p.adj < p] <- "UP"
              de$diffexpressed[de$log2FC < -log_thresh & de$p.adj < p] <- "DOWN"
              de$diffexpressed <- as.character(de$diffexpressed)
              de$delabel <- NA
              de$delabel[de$diffexpressed != "NO"] <- de$Gene[de$diffexpressed != "NO"]
              de <- de %>% arrange(-log2FC, p_val_adj)
              de$contrast <- contrast.str

              write.csv(de, file.path(out.gene.lists, fn))
            }

            if (save.filtered){
              filt.de <- de %>% filter(diffexpressed %in% c('UP','DOWN'))
              fn.filt <- paste0('filt.', base.fn, '.csv')
              write.csv(filt.de, file.path(filt.out.gene.lists, fn.filt))
            }

            if (plot.volc){
              p1 <- volcano.plot(de, p, log_thresh, title)
              save.figs(p1, out.de.volcano, base.fn, width=width, height=height)
            }

            if (plot.heat){
  if (cell == 'all.cells'){ ds.pt <- 10;  ndeg <- 10
  } else {                  ds.pt <- 200; ndeg <- 15 }

  p2 <- tryCatch({
    heatmap.plot(de, sub, title, plot.tag = tag,
                 ndeg_top_bottom = ndeg, p.thresh = p,
                 ds.pt = ds.pt)
  }, error = function(e) {
    warning("heatmap.plot error for ", base.fn, ": ", e$message)
    NA
  })

  # FIX: check class — pheatmap returns a list, NA returns logical
  if (is.list(p2)) {
    save.figs(p2, out.de.heatmap, base.fn, width=10, height=5)
  } else {
    print(paste('Too few genes for heatmap:', base.fn))
  }
}

          } # end pair loop
        } # end cell loop
      } # end trt loop
    } # end tissue loop
} # end tag loop  ← moved inside time loop so tag is defined before time.sub logic
} # end time loop
} # end log_thresh loop
}) # end system.time
# Run outside heatmap.plot to identify the correct assay
cat("Assay names:", paste(Assays(sub), collapse=", "), "\n")
cat("RNA nrow:", nrow(sub[["RNA"]]), "\n")
cat("First 5 RNA rownames:", paste(head(rownames(sub[["RNA"]]), 5), collapse=", "), "\n")

# Check if genes are in a different assay
for (assay_name in Assays(sub)) {
  n_match <- sum(use.genes %in% rownames(sub[[assay_name]]))
  cat(assay_name, "matches:", n_match, "/", length(use.genes), "\n")

```
```{r}
# Annotation color generator
ann.colors <- function(annotation_col, sub) {
  color_map <- list(
    PFS.6month        = c("NoPrg6M"       = "coral",       "Prg6M"         = "slateblue"),
    timepoint         = c("Baseline"       = "grey90",      "Week4"         = "hotpink1",  "Week12.or.16" = "purple"),
    treatment         = c("combo"          = "grey60",      "nivo"          = "cyan",      "rela"         = "magenta"),
    general_cell_type = c("CD8plus.T.cell" = "yellowgreen", "Macrophages"   = "violet",    "Treg"         = "#FDC086")
  )

  present_tags <- colnames(annotation_col)
  matched      <- intersect(present_tags, names(color_map))
  unmatched    <- setdiff(present_tags, names(color_map))

  if (length(unmatched) > 0)
    warning("No color definition for tag(s), pheatmap will auto-color: ",
            paste(unmatched, collapse = ", "))

  result <- list()
  for (tag in matched) {
    actual_levels <- unique(as.character(annotation_col[[tag]]))
    actual_levels <- actual_levels[!is.na(actual_levels)]
    defined       <- color_map[[tag]]

    missing_levels <- setdiff(actual_levels, names(defined))
    if (length(missing_levels) > 0)
      warning("Tag '", tag, "' levels in data not in color definition: ",
              paste(missing_levels, collapse = ", "))

    # FIX: only keep color entries for levels that actually appear in the data
    result[[tag]] <- defined[intersect(names(defined), actual_levels)]
  }

  result
}

# pheatmap wrapper
pheatmap.wrapper <- function(lognorm.dat, annotation_col, ann_colors, title,
                             cluster_rows = TRUE,
                             cluster_cols = FALSE) {
  pheatmap::pheatmap(
    lognorm.dat,
    main              = title,
    scale             = "none",
    annotation_col    = annotation_col,
    annotation_colors = ann_colors,
    clustering_method = "complete",
    cluster_rows      = cluster_rows,
    cluster_cols      = cluster_cols,
    color             = colorRampPalette(c("navy", "white", "firebrick3"))(100),
    show_colnames     = FALSE,
    show_rownames     = TRUE
  )
}

# Manual row-scaling
row.scale <- function(mat) {
  t(apply(mat, 1, function(x) {
    s <- sd(x, na.rm = TRUE)
    if (is.na(s) || s == 0) return(rep(0, length(x)))
    (x - mean(x, na.rm = TRUE)) / s
  }))
}

# Main heatmap function
heatmap.plot <- function(de, sub, title, ds.pt = 100,
                         ndeg_top_bottom = 15,
                         p.thresh        = 0.05,
                         plot.tag        = c('PFS.6month', 'timepoint', 'treatment', 'general_cell_type')) {

  DefaultAssay(sub) <- "RNA"
  Idents(sub)       <- 'general_cell_type'
  suppressWarnings({ sub <- subset(sub, downsample = ds.pt) })

  missing_tags <- setdiff(plot.tag, colnames(sub@meta.data))
  if (length(missing_tags) > 0)
    warning("plot.tag(s) not found in metadata and skipped: ",
            paste(missing_tags, collapse = ", "))

  # FIX: keep as character, not factor — pheatmap handles its own factor conversion
  annotation_col <- sub@meta.data %>%
    select(any_of(plot.tag)) %>%
    mutate(across(everything(), as.character)) %>%
    as.data.frame()

  # FIX: build ann_colors from actual annotation_col after subsetting
  ann_colors <- ann.colors(annotation_col, sub)

  top    <- de %>% filter(p_val_adj < p.thresh, avg_log2FC >  0) %>% top_n(ndeg_top_bottom, wt =  avg_log2FC)
  bottom <- de %>% filter(p_val_adj < p.thresh, avg_log2FC <  0) %>% top_n(ndeg_top_bottom, wt = -avg_log2FC)

  use.genes <- unique(rbind(top, bottom)$Gene)

  if (length(use.genes) <= 1) {
    warning("Fewer than 2 DEGs passed filters for: ", title)
    return(NA)
  }

  available.genes <- intersect(use.genes, rownames(sub[["RNA"]]))
  if (length(available.genes) < length(use.genes))
    warning("Some genes not found in RNA assay and dropped: ",
            paste(setdiff(use.genes, available.genes), collapse = ", "))

  lognorm.dat <- as.matrix(
    GetAssayData(sub, assay = "RNA", slot = "data")[available.genes,
                                                    row.names(annotation_col),
                                                    drop = FALSE]
  )

  # Guard 1: Inf/-Inf → NA
  lognorm.dat[is.infinite(lognorm.dat)] <- NA

  # Guard 2: drop all-NA rows
  all.na.rows <- apply(lognorm.dat, 1, function(x) all(is.na(x)))
  if (any(all.na.rows)) {
    warning("Dropping ", sum(all.na.rows), " gene(s) with all-NA expression")
    lognorm.dat <- lognorm.dat[!all.na.rows, , drop = FALSE]
  }

  # Guard 3: impute remaining NAs with row mean
  lognorm.dat <- t(apply(lognorm.dat, 1, function(x) {
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    x
  }))

  # Guard 4: drop zero/NA variance columns
  col.var       <- apply(lognorm.dat, 2, var)
  zero.var.cols <- is.na(col.var) | col.var == 0
  if (any(zero.var.cols)) {
    warning("Dropping ", sum(zero.var.cols), " cell(s) with zero/NA variance")
    lognorm.dat    <- lognorm.dat[, !zero.var.cols, drop = FALSE]
    annotation_col <- annotation_col[!zero.var.cols, , drop = FALSE]
  }

  # Guard 5: manual row scaling
  lognorm.dat <- row.scale(lognorm.dat)

  # Guard 6: final dimension and value check
  if (nrow(lognorm.dat) <= 1 || ncol(lognorm.dat) <= 1) {
    warning("Too few genes/cells remain after cleaning for: ", title)
    return(NA)
  }

  if (any(is.nan(lognorm.dat)) || any(is.infinite(lognorm.dat))) {
    warning("NaN/Inf remain after cleaning, skipping heatmap for: ", title)
    return(NA)
  }

  pheatmap.wrapper(lognorm.dat, annotation_col, ann_colors, title)
}
```
```{r}
# Temporary diagnostic wrapper — run this instead of heatmap.plot once to see what's happening
heatmap.plot.debug <- function(de, sub, title, ds.pt = 100,
                               ndeg_top_bottom = 15,
                               p.thresh        = 0.05,
                               plot.tag        = c('PFS.6month', 'timepoint', 'treatment', 'general_cell_type')) {

  DefaultAssay(sub) <- "RNA"
  Idents(sub)       <- 'general_cell_type'
  suppressWarnings({ sub <- subset(sub, downsample = ds.pt) })

  annotation_col <- sub@meta.data %>%
    select(any_of(plot.tag)) %>%
    mutate(across(everything(), as.character)) %>%
    as.data.frame()

  cat("\n=== annotation_col columns ===\n")
  print(colnames(annotation_col))

  cat("\n=== unique values per column ===\n")
  for (col in colnames(annotation_col)) {
    cat(col, ":", paste(unique(annotation_col[[col]]), collapse = ", "), "\n")
  }

  cat("\n=== ann_colors keys and levels ===\n")
  ac <- ann.colors(annotation_col, sub)
  for (nm in names(ac)) {
    cat(nm, ":", paste(names(ac[[nm]]), collapse = ", "), "\n")
  }

  cat("\n=== colnames match names(ann_colors)? ===\n")
  cat("annotation_col cols:", paste(colnames(annotation_col), collapse=", "), "\n")
  cat("ann_colors names:   ", paste(names(ac), collapse=", "), "\n")
  cat("Match:", identical(sort(colnames(annotation_col)), sort(names(ac))), "\n")
}

# Call on the specific tag/sub combination that's failing
heatmap.plot.debug(de, sub, title, plot.tag = tag)
```
```{r}
# Annotation color generator
ann.colors <- function(annotation_col) {  # removed 'sub' — not needed
  color_map <- list(
    PFS.6month        = c("NoPrg6M"       = "coral",       "Prg6M"         = "slateblue"),
    timepoint         = c("Baseline"       = "grey90",      "Week4"         = "hotpink1",  "Week12.or.16" = "purple"),
    treatment         = c("combo"          = "grey60",      "nivo"          = "cyan",      "rela"         = "magenta"),
    general_cell_type = c("CD8plus.T.cell" = "yellowgreen", "Macrophages"   = "violet",    "Treg"         = "#FDC086")
  )

  result <- list()
  for (col in colnames(annotation_col)) {
    if (!col %in% names(color_map)) next
    actual_levels <- unique(as.character(annotation_col[[col]]))
    actual_levels <- actual_levels[!is.na(actual_levels)]
    defined       <- color_map[[col]]
    # Only keep colors for levels present in data; warn on any missing
    missing_levels <- setdiff(actual_levels, names(defined))
    if (length(missing_levels) > 0)
      warning("Tag '", col, "' has uncolored levels: ",
              paste(missing_levels, collapse = ", "))
    result[[col]] <- defined[names(defined) %in% actual_levels]
  }
  result
}

# pheatmap wrapper
pheatmap.wrapper <- function(lognorm.dat, annotation_col, ann_colors, title,
                             cluster_rows = TRUE,
                             cluster_cols = FALSE) {

  # Convert each annotation column to factor with levels = exactly the
  # keys present in ann_colors for that column (pheatmap requirement)
  annotation_col <- as.data.frame(annotation_col)  # ensure plain data.frame
  for (col in colnames(annotation_col)) {
    vals <- as.character(annotation_col[[col]])
    lvls <- if (col %in% names(ann_colors)) {
      names(ann_colors[[col]])[names(ann_colors[[col]]) %in% unique(vals)]
    } else {
      sort(unique(vals))
    }
    annotation_col[[col]] <- factor(vals, levels = lvls)
  }

  # Pre-flight: verify keys match exactly before pheatmap sees them
  ann_colors <- ann_colors[names(ann_colors) %in% colnames(annotation_col)]

  tryCatch({
    pheatmap::pheatmap(
      lognorm.dat,
      main              = title,
      scale             = "none",
      annotation_col    = annotation_col,
      annotation_colors = ann_colors,
      clustering_method = "complete",
      cluster_rows      = cluster_rows,
      cluster_cols      = cluster_cols,
      color             = colorRampPalette(c("navy", "white", "firebrick3"))(100),
      show_colnames     = FALSE,
      show_rownames     = TRUE
    )
  }, error = function(e) {
    warning("pheatmap failed with annotation_colors (", e$message, "). Retrying without colors.")
    pheatmap::pheatmap(
      lognorm.dat,
      main              = title,
      scale             = "none",
      annotation_col    = annotation_col,
      clustering_method = "complete",
      cluster_rows      = cluster_rows,
      cluster_cols      = cluster_cols,
      color             = colorRampPalette(c("navy", "white", "firebrick3"))(100),
      show_colnames     = FALSE,
      show_rownames     = TRUE
    )
  })
}

# Manual row-scaling
row.scale <- function(mat) {
  t(apply(mat, 1, function(x) {
    s <- sd(x, na.rm = TRUE)
    if (is.na(s) || s == 0) return(rep(0, length(x)))
    (x - mean(x, na.rm = TRUE)) / s
  }))
}

# Main heatmap function
heatmap.plot <- function(de, sub, title, ds.pt = 100,
                         ndeg_top_bottom = 15,
                         p.thresh        = 0.05,
                         plot.tag        = c('PFS.6month', 'timepoint', 'treatment', 'general_cell_type')) {

  DefaultAssay(sub) <- "RNA"
  Idents(sub)       <- 'general_cell_type'
  suppressWarnings({ sub <- subset(sub, downsample = ds.pt) })

  missing_tags <- setdiff(plot.tag, colnames(sub@meta.data))
  if (length(missing_tags) > 0)
    warning("plot.tag(s) not found in metadata and skipped: ",
            paste(missing_tags, collapse = ", "))

  annotation_col <- sub@meta.data %>%
    select(any_of(plot.tag)) %>%
    mutate(across(everything(), as.character)) %>%
    as.data.frame()

  # Build colors from the actual annotation_col that will be passed to pheatmap
  ann_colors <- ann.colors(annotation_col)

  top    <- de %>% filter(p_val_adj < p.thresh, avg_log2FC >  0) %>% top_n(ndeg_top_bottom, wt =  avg_log2FC)
  bottom <- de %>% filter(p_val_adj < p.thresh, avg_log2FC <  0) %>% top_n(ndeg_top_bottom, wt = -avg_log2FC)

  use.genes <- unique(rbind(top, bottom)$Gene)

  if (length(use.genes) <= 1) {
    warning("Fewer than 2 DEGs passed filters for: ", title)
    return(NA)
  }

  available.genes <- intersect(use.genes, rownames(sub[["RNA"]]))
  if (length(available.genes) < length(use.genes))
    warning("Some genes not found in RNA assay and dropped: ",
            paste(setdiff(use.genes, available.genes), collapse = ", "))

  lognorm.dat <- as.matrix(
    GetAssayData(sub, assay = "RNA", slot = "data")[available.genes,
                                                    row.names(annotation_col),
                                                    drop = FALSE]
  )

  # Guard 1: Inf/-Inf → NA
  lognorm.dat[is.infinite(lognorm.dat)] <- NA

  # Guard 2: drop all-NA rows
  all.na.rows <- apply(lognorm.dat, 1, function(x) all(is.na(x)))
  if (any(all.na.rows)) {
    warning("Dropping ", sum(all.na.rows), " gene(s) with all-NA expression")
    lognorm.dat <- lognorm.dat[!all.na.rows, , drop = FALSE]
  }

  # Guard 3: impute remaining NAs with row mean
  lognorm.dat <- t(apply(lognorm.dat, 1, function(x) {
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    x
  }))

  # Guard 4: drop zero/NA variance columns and sync annotation_col
  col.var       <- apply(lognorm.dat, 2, var)
  zero.var.cols <- is.na(col.var) | col.var == 0
  if (any(zero.var.cols)) {
    warning("Dropping ", sum(zero.var.cols), " cell(s) with zero/NA variance")
    lognorm.dat    <- lognorm.dat[, !zero.var.cols, drop = FALSE]
    annotation_col <- annotation_col[!zero.var.cols, , drop = FALSE]
  }

  # Guard 5: manual row scaling
  lognorm.dat <- row.scale(lognorm.dat)

  # Guard 6: final dimension and value check
  if (nrow(lognorm.dat) <= 1 || ncol(lognorm.dat) <= 1) {
    warning("Too few genes/cells remain after cleaning for: ", title)
    return(NA)
  }

  if (any(is.nan(lognorm.dat)) || any(is.infinite(lognorm.dat))) {
    warning("NaN/Inf remain after cleaning, skipping heatmap for: ", title)
    return(NA)
  }

  # Rebuild ann_colors after guard 4 may have changed annotation_col
  ann_colors <- ann.colors(annotation_col)

  pheatmap.wrapper(lognorm.dat, annotation_col, ann_colors, title)
}
```
```{r}
heatmap.plot <- function(de, sub, title, ds.pt = 100,
                         ndeg_top_bottom = 15,
                         p.thresh        = 0.05,
                         plot.tag        = c('PFS.6month', 'timepoint', 'treatment', 'general_cell_type')) {

  # --- Diagnostic: check de structure on entry ---
  cat("\n--- heatmap.plot diagnostics ---\n")
  cat("de cols:", paste(colnames(de), collapse=", "), "\n")
  cat("nrow(de):", nrow(de), "\n")
  cat("p_val_adj range:", range(de$p_val_adj, na.rm=TRUE), "\n")
  cat("avg_log2FC range:", range(de$avg_log2FC, na.rm=TRUE), "\n")
  cat("'Gene' col exists:", "Gene" %in% colnames(de), "\n")
  cat("p.thresh used:", p.thresh, "\n")

  # FIX: avg_log2FC in saved CSV may be stored under 'log2FC' after de$log2FC <- de[[log.name]]
  # Harmonize: use whichever FC column exists
  if (!"avg_log2FC" %in% colnames(de) && "log2FC" %in% colnames(de)) {
    de$avg_log2FC <- de$log2FC
    cat("Using 'log2FC' as avg_log2FC\n")
  }

  # FIX: p_val_adj in saved CSV may be stored under 'p.adj'
  if (!"p_val_adj" %in% colnames(de) && "p.adj" %in% colnames(de)) {
    de$p_val_adj <- de$p.adj
    cat("Using 'p.adj' as p_val_adj\n")
  }

  top    <- de %>% filter(p_val_adj < p.thresh, avg_log2FC >  0) %>% top_n(ndeg_top_bottom, wt =  avg_log2FC)
  bottom <- de %>% filter(p_val_adj < p.thresh, avg_log2FC <  0) %>% top_n(ndeg_top_bottom, wt = -avg_log2FC)

  cat("top DEGs:", nrow(top), "| bottom DEGs:", nrow(bottom), "\n")

  use.genes <- unique(rbind(top, bottom)$Gene)
  cat("use.genes:", length(use.genes), "\n")

  if (length(use.genes) <= 1) {
    warning("Fewer than 2 DEGs passed filters for: ", title)
    return(NA)
  }

  DefaultAssay(sub) <- "RNA"
  Idents(sub)       <- 'general_cell_type'
  suppressWarnings({ sub <- subset(sub, downsample = ds.pt) })

  missing_tags <- setdiff(plot.tag, colnames(sub@meta.data))
  if (length(missing_tags) > 0)
    warning("plot.tag(s) not found in metadata and skipped: ",
            paste(missing_tags, collapse = ", "))

  annotation_col <- as.data.frame(
    sub@meta.data[, intersect(plot.tag, colnames(sub@meta.data)), drop = FALSE]
  )
  annotation_col[] <- lapply(annotation_col, as.character)

  available.genes <- intersect(use.genes, rownames(sub[["RNA"]]))
  if (length(available.genes) < length(use.genes))
    warning("Some genes not found in RNA assay and dropped: ",
            paste(setdiff(use.genes, available.genes), collapse = ", "))

  if (length(available.genes) <= 1) {
    warning("Fewer than 2 available genes in RNA assay for: ", title)
    return(NA)
  }

  lognorm.dat <- as.matrix(
    GetAssayData(sub, assay = "RNA", slot = "data")[available.genes,
                                                    rownames(annotation_col),
                                                    drop = FALSE]
  )

  # Guard 1: Inf/-Inf → NA
  lognorm.dat[is.infinite(lognorm.dat)] <- NA

  # Guard 2: drop all-NA rows
  all.na.rows <- apply(lognorm.dat, 1, function(x) all(is.na(x)))
  if (any(all.na.rows)) {
    warning("Dropping ", sum(all.na.rows), " gene(s) with all-NA expression")
    lognorm.dat <- lognorm.dat[!all.na.rows, , drop = FALSE]
  }

  # Guard 3: impute remaining NAs with row mean
  lognorm.dat <- t(apply(lognorm.dat, 1, function(x) {
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    x
  }))

  # Guard 4: drop zero/NA variance columns — keep annotation_col in sync
  col.var       <- apply(lognorm.dat, 2, var)
  zero.var.cols <- is.na(col.var) | col.var == 0
  if (any(zero.var.cols)) {
    warning("Dropping ", sum(zero.var.cols), " cell(s) with zero/NA variance")
    lognorm.dat    <- lognorm.dat[, !zero.var.cols, drop = FALSE]
    annotation_col <- as.data.frame(annotation_col[!zero.var.cols, , drop = FALSE])
  }

  # Guard 5: manual row scaling
  lognorm.dat <- row.scale(lognorm.dat)

  # Guard 6: final dimension and value check
  if (nrow(lognorm.dat) <= 1 || ncol(lognorm.dat) <= 1) {
    warning("Too few genes/cells remain after cleaning for: ", title)
    return(NA)
  }

  if (any(is.nan(lognorm.dat)) || any(is.infinite(lognorm.dat))) {
    warning("NaN/Inf remain after cleaning, skipping heatmap for: ", title)
    return(NA)
  }

  ann_colors <- ann.colors(annotation_col)

  pheatmap.wrapper(lognorm.dat, annotation_col, ann_colors, title)
}

```
```{r}

files.pfs.pbmc = list.files(path = "DEG_contrasts_PBMC_MAST_v3/filtered_csv/", pattern="*PFS*.*\\.csv$", full.names = TRUE)
combined_data.pfs.pbmc <- files.pfs.pbmc %>%
    set_names() %>% # Uses filenames as identifiers
    map_dfr(read_csv, .id = "source_file") # Adds 'source_file' column
```
Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
